---
title: "Lab 5 In class"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## NPZD Model 

Nutrients, Plankton, Zooplankton, and Detritus (NPZD) Model.

### Drivers

First we start with a function will we use to simulate how solar energy from the sun changes over the year. Solar energy the environmental "driver" of our model. The function is used in our differential equation below.

```{r}
light_function <- function(x){
  t <- x
  0.5 * (540 + 440 * sin(2 * pi * t / 365 - 1.4))
}

curve(light_function(x), 
      from = 1, to = 730, 
      ylab = "umol/m2/s", 
      xlab = "# days", 
      main = "Light input")
```
### Model 

Here is our differential equation model based on our conceptual model.  

```{r}
NPZD_model <- function(t, states, parms){
  
  #unpack states
  PHYTO <- states[1]
  ZOO <- states[2]
  DET <- states[3]
  DIN <- states[4]
  
  #unpack parms
  maxUptake <- parms[1]
  kspar <- parms[2] #Sunlight Limiting
  ksdin <- parms[3] #Nitrogen limiting
  maxGrazing <- parms[4]
  ksphyto <- parms[5]
  pFeces <- parms[6] #proportion of waste to feces
  excretionRate <- parms[7]
  mortalityRate <- parms[8]
  mineralizationRate <- parms[9]
  Chl_Nratio <- parms[10]
  
  #Calculate the environment (note that this depends on time)
  #50% of light is PAR
  PAR <- 0.5 * (540 + 440 * sin(2 * pi * t / 365 - 1.4)) 
  
  #calculate fluxes
  fDIN <- DIN / (DIN + ksdin)
  fLight <- PAR / (PAR + kspar)
  uptake <- maxUptake * PHYTO * fDIN * fLight #min(c(fDIN, fLight))
  
  fPHYTO <- PHYTO / (PHYTO + ksphyto)
  grazing <- maxGrazing * ZOO * fPHYTO
  
  feces_production <- grazing * pFeces
  
  excretion <- excretionRate * ZOO
  
  zoo_mortality <- mortalityRate * ZOO^2
  
  mineralization <- mineralizationRate * DET
  
  #calculate extra outputs
  Chlorophyll <- PHYTO * Chl_Nratio
  
  #calculate derivatives
  dPHYTO_dt <- uptake - grazing
  dZOO_dt <- grazing - feces_production - zoo_mortality - excretion
  dDETRITUS_dt <- zoo_mortality + feces_production - mineralization
  dDIN_dt <- excretion + mineralization - uptake

 #return the list of derivatives plus any other variables that you are interested in
  
  return(list(c(dPHYTO_dt = dPHYTO_dt, # the rates of change
                dZOO_dt = dZOO_dt,
                dDETRITUS_dt = dDETRITUS_dt,
                dDIN_dt = dDIN_dt),
              # other output variables
              c(Chlorophyll = Chlorophyll, 
                PAR = PAR)))         
}

```

### Parameters

```{r}
parms <- c(
  1.0,  #maxUptake INSERT UNITS
  120,  #kspar INSERT UNITS
  0.5,  #ksdin INSERT UNITS
  1.0,  #maxGrazing INSERT UNITS
  1.0,    #ksphyto INSERT UNITS
  0.3,  #pFaeces INSERT UNITS
  0.1,  #excretionRate INSERT UNITS
  0.4,  #mortalityRate INSERT UNITS
  0.1,  #mineralizationRate INSERT UNITS
  1.0)    #Chl_Nratio INSERT UNITS
```

### Initial conditions

```{r}
yini <- c(
  PHYTO = 1, #INSERT UNITS
  ZOO = 0.8,  #INSERT UNITS
  DETRITUS = 1,  #INSERT UNITS
  DIN = 9)  #INSERT UNITS
```

### Solve model

```{r}
#Use DeSolve to integrate 
library(deSolve)

simulation_time <- 365 * 2
dt <- 1
times <- seq(1, simulation_time, by = dt)

output <- ode(y = yini, times = times, func = NPZD_model, params = parms, method = "ode45")
```

### Visualize output

```{r}
#Plot the model output
#First convert output to a data frame
output <- as.data.frame(output)
par(mfrow = c(2, 3))
plot(times, output$PAR, xlab = "time (days)", ylab = "umol/m2/s", main = "PAR", type = "l")
plot(times, output$PHYTO, xlab="time (days)", ylab ="mmolN/m3", main = "Phytoplankton", type = "l")
plot(times, output$Chlorophyll, xlab = "time (days)", ylab = "ug/l", main = "Chlorophyll", type = "l")
plot(times, output$ZOO, xlab = "time (days)", ylab = "mmolN/m3", main = "Zooplankton", type = "l")
plot(times, output$DIN, xlab = "time (days)", ylab = "mmolN/m3",main="DIN", type = "l")
plot(times, output$DETRITUS, xlab = "time (days)", ylab = "mmolN/m3", main = "DETRITUS", type = "l")
```
