---
title: "Ecosystem Dynamics Lab 5"
output:
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Goals:
-	Practice adding a new flux to a differential equation model
-	Apply the simple NPZD model to a new science question
-	Practice analyzing model behavior

We will apply our NPZD model (S&H Section 2.9.1) to investigate the influence of nitrogen pollution on our ecosystem. The current model has no N inputs or outputs, so we need to add the fluxes.  

# Step 1: Conceptual Model

**Question 1:**  

Draw a conceptual diagram (building on S&H figure 2.20 diagram) that adds two new fluxes:  Dissolved inorganic nitrogen (DIN) input from agricultural run-off and DIN output from a stream that drains the lake.

**Answer 1:**
![](INSERT PATH TO DIAGRAM)

Modify your model so that the two new fluxes are now included.  The modifications will not require adding any new stocks, but you will have to add two new parameters. 

#Step 2: Code Model

```{r}
#INSERT MODEL HERE

```

#Step 3: Parameterize Model

We are going to assume that the DIN output is a first-order process controlled by the DIN in the lake. The DIN output rate parameter will be the turnover rate of water in the lake (1 / residence time). Assume that the lake has a 720-day residence time.

We will assume that the DIN input is a constant equal to 0.0012 mmol N m-3 day-1.

```{r}
#INSERT PARAMETERS HERE
```

#Step 4: Initialize Model

```{r}
#INSERT INITIAL CONDITIONS HERE
yini <- c(
  PHYTO = 1, #mmolN m-3
  ZOO = 0.8, #mmolN m-3
  DETRITUS = 1, #mmolN m-3
  DIN = 9) #mmolN m-3
```


#Step 5: Verify Model

**Question 2:** 

What is the mid-summer stock of phytoplankton after 10 years? Are the dynamics reasonable? Include a plot of the time series from the simulation.  

**Answer 2:** 

```{r}

```

#Step 6: Apply model

To meet regulations, the lake needs the summer phytoplankton population to not exceed 2 mmol N m^-3 within the next 10 years.

**Question 3:** 

What is the smallest reduction in agriculture run-off that will achieve this goal? Give your answer as a percentage reduction from the baseline simulation reported in question 2

**Answer 3:**

```{r}


```

**Question 4:**

Provide a plot of phytoplankton over the 10-year period with and without the reduction in run-off.  Ensure the figure includes a legend, properly labeled axes, and a title.

**Answer 4:**

```{r}


```

#Step 7: Improve the model

**Question 5:**  

You are an ecosystem scientist tasked with improving the model by better parameterizing the model that it models phytoplankton more realistically.  To address this task, you are designing a field-sampling program to measure the model's parameters.  Which parameter should get the most attention in the field program and why?  Provide a quantitative assessment, including a description of the approach you used to make the assessment.

Focus on the following 4 parameters: maxNuptake, maxGrazing, kspar, ksdin   

**Answer 5:**

```{r}


```




