---
title: "Ecosystem Dynamics Lab 8"
output:
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk $set(echo = TRUE)
```

# Goals: 

-	Construct a model from peer-reviewed literature
-	Learn how to design a modeling study
-	Use R to solve differential equations
-	Create clean graphs to communicate model experiments
-	Explore an alternative decomposition model to the classic first-order decay model
- Build independence in your modeling abilities

# Step 1: Build model

Construct the enzyme model described by Allison et al. (2010).  Use the modeling framework developed in class to build the differential equation model. 

```{r}
enzyme_catalysis_model <- function(t, state, params){
  #unpack states
  SOC <- state[1]
  DOC <- state[2]
  ENZ <- state[3]
  MIC <- state[4]
  
  #unpack parameters
    # Parameters relevant to dSOC/dt:
    SOCin <- params[1] # input of SOC
    rDeath <- params[2] # rate of Death
    MIC2SOC <- params[3] #Proportion death of MIC to SOC
    Ea_decomp <- params[4] #Velocity of decomposition
    Km_slope_decomp <- params[5] 
    temp <- params[6]
    Km0_decomp <- params[7]
    
    # Parameters relevant to dMIC/dt:
    Ea_uptake <- params[8]
    Km_slope_uptake <- params[9]
    Km0_uptake <- params[10]
    #rDeath pre defined
    r_EnzProd <- params[11]
    
    #Parameters relevant to dDOC/dt:
    DOCin <- params[12]
    #rDeath, MICtoSOC, Decomp predefined
    r_EnzLoss <- params[13]
    
    
    #CUE parameters:
    CUE_slope <- params[14]
    CUE0 <- params[15]
    Vmax_uptake0 <- params[16]
    Vmax_decomp0 <- params[17]
    gasConst <- params[18]
    
    
  #Calculate intermediate parameters:
    Vmax_decomp <- Vmax_decomp0 * exp(-Ea_decomp/(gasConst*(temp +273)))
    Vmax_uptake <- Vmax_uptake0 * exp(-Ea_uptake/(gasConst*(temp + 273)))
    Km_decomp <- Km_slope_decomp * temp + Km0_decomp
    Km_uptake <- Km_slope_uptake * temp + Km0_uptake
    CUE <- CUE_slope * temp + CUE0
    Death <- rDeath * MIC
    Assim <- Vmax_uptake * MIC * (DOC/(Km_uptake + DOC))
    EProd <- r_EnzProd * MIC
    Decomp <- Vmax_decomp * ENZ * (SOC/(Km_decomp + SOC))
    ELoss <- r_EnzLoss * ENZ
    
    CO2 <- Assim * (1-CUE)
   
   
    
  #Diffeqs:
  dMIC <- Assim * CUE - Death - EProd
  
  dSOC <- SOCin + (Death*MIC2SOC) - Decomp
  
  dDOC <- DOCin + (Death*(1-MIC2SOC)) + Decomp + ELoss - Assim
  
  dENZ <- EProd - ELoss
  
  
  return(list(c(dSOC = dSOC,
                dDOC = dDOC,
                dENZ = dENZ,
                dMIC = dMIC),
              c(CO2 = CO2)))
}


```

# Step 2: Parameterize model

Design the model experiments by setting up the different parameter sets required to run the experiments shown in column (a) of Figure 2

- control (no warming).  
- warm + const CUE. 
- warm + vary CUE. 
- warm + acclim CUE. 

Remember that you will need to do a spin-up simulation to determine the initial conditions (`yini`).  The same initial conditions will be used for each of the four model experiments above.  

One of the challenges of the assignment is figuring out how the authors carried out their model experiments.

###Parameters
```{r}
  params <- c(
    .0005, #Input SOC
    .0002, #Rate of Death
    .5, #MIC2SOC
    47, #Ea-decomp
    5, #Km Slope decomp
    20, #temp
    500, #Km0
    47, #Ea Uptake
    .01, #kmslope uptake
    .1, #km0 uptake
    .000005, #Rate of enzyme production
    .0005, #DOC input
    .001, #Rate of enzyme loss
    -.016, # CUE Slope
    .63, #CUE 0
    100000000, #Vmax uptake
    100000000, #Vmax Decomp
    .008314 #Gas constant
)
  
```

###Spin up initial  Conditions
```{r}
yini_spinup <- c(
  SOC = 100, #100
  DOC = .5, #.5
  ENZ = .01, #.01
  MIC = .5 #.5
)
```


###Solve Initial model

```{r}
library(deSolve)

simulation_time <- 876000
times <- seq(1, simulation_time, by = 24)
output <- ode(y = yini_spinup, times = times, func = enzyme_catalysis_model, parms = params, method = "lsoda")
yini_calculated <- tail(output, 1)
yini <- c(SOC = yini_calculated[1], DOC = yini_calculated[2], ENZ = yini_calculated[3], MIC = yini_calculated[4])
```


##Experiments

###Control
```{r}
simulation_time <- 262800
times <- seq(1, simulation_time, by = 24)
control <- ode(y = yini, times = times, func = enzyme_catalysis_model, parms = params, method = "lsoda")
control <- as.data.frame(control)
```

###Warming + const. CUE
```{r}
params[6] = 25
params[15] = .31
params[14] = 0

warming <- ode(y = yini, times = times, func = enzyme_catalysis_model, parms = params, method = "lsoda")
warming <- as.data.frame(warming)
```

###Warming + vary CUE
```{r}
params[6] = 25
params[15] = .63
params[14] = -.016

warming_VCUE <- ode(y = yini, times = times, func = enzyme_catalysis_model, parms = params, method = "lsoda")
warming_VCUE <- as.data.frame(warming_VCUE)
```

###Warming + acclim CUE
```{r}
params[6] = 25
params[15] = .27
params[14] = 0

warming_ACUE <- ode(y = yini, times = times, func = enzyme_catalysis_model, parms = params, method = "lsoda")
warming_ACUE <- as.data.frame(warming_ACUE)
```


# Step 3: Run model

Run the simulations necessary to reproduce the results in column a of Figure 2 in Allison et al. (2010)  Use deSolve to solve your model.

Tips for solving the model

- In the ode function, change `method = ode23` to `method = lsoda`
- Use the the following code to get the state values at the last time step from the spin-up simulation,  Use these values as the `yini` for the experiments

```{r eval=FALSE}
tail(out,1) #where 'out' is the output of the ODE solver
```

- Include 'CO2' as one of the extra outputs in the differential equation function because it is not a model state, but you want to plot it.  Please take a look at the CHL example in the NPZD lab.
- Include temperature as a parameter (though I like to think of it as a driver or forcing)
- In the spin-up, use a simulation time of 100 years (in hours) and a dt of 24 hrs (one day)
- In the experiments, use a simulation time of 30 years (in hours) and a dt of 24 hrs (one day)

#Plots
```{r, fig.width=10, fig.height=8}
library(ggplot2)
library("ggpubr")
control$time <- control$time/8760
CO2_p <- ggplot() + 
geom_line(mapping= aes(x= (control$time),y=control$CO2.MIC, color = "Control")) +
geom_line(mapping= aes(x= (control$time),y=warming$CO2.MIC, color = "Warming + Constant CUE")) +
geom_line(mapping= aes(x=(control$time),y=warming_ACUE$CO2.MIC, color = "Warming + Acclim. CUE")) +
geom_line(mapping= aes(x=(control$time),y=warming_VCUE$CO2.MIC, color = "Warming + Varying CUE")) +
  labs(x = "Time (yrs)", y = "CO2")

DOC_p <- ggplot() + 
geom_line(mapping= aes(x=control$time,y=control$DOC, color = "Control")) +
geom_line(mapping= aes(x=control$time,y=warming$DOC, color = "Warming + Constant CUE")) +
geom_line(mapping= aes(x=control$time,y=warming_ACUE$DOC, color = "Warming + Acclim. CUE")) +
geom_line(mapping= aes(x=control$time,y=warming_VCUE$DOC, color = "Warming + Varying CUE")) +
    labs(x = "Time (yrs)", y = "DOC")

SOC_p <- ggplot() + 
geom_line(mapping= aes(x=control$time,y=control$SOC, color = "Control")) +
geom_line(mapping= aes(x=control$time,y=warming$SOC, color = "Warming + Constant CUE")) +
geom_line(mapping= aes(x=control$time,y=warming_ACUE$SOC, color = "Warming + Acclim. CUE")) +
geom_line(mapping= aes(x=control$time,y=warming_VCUE$SOC, color = "Warming + Varying CUE")) +
    labs(x = "Time (yrs)", y = "SOC")


ENZ_p <- ggplot() + 
geom_line(mapping= aes(x=control$time,y=control$ENZ, color = "Control")) +
geom_line(mapping= aes(x=control$time,y=warming$ENZ, color = "Warming + Constant CUE")) +
geom_line(mapping= aes(x=control$time,y=warming_ACUE$ENZ, color = "Warming + Acclim. CUE")) +
geom_line(mapping= aes(x=control$time,y=warming_VCUE$ENZ, color = "Warming + Varying CUE")) +
    labs(x = "Time (yrs)", y = "ENZ")


MIC_p <- ggplot() + 
geom_line(mapping= aes(x=control$time,y=control$MIC, color = "Control")) +
geom_line(mapping= aes(x=control$time,y=warming$MIC, color = "Warming + Constant CUE")) +
geom_line(mapping= aes(x=control$time,y=warming_ACUE$MIC, color = "Warming + Acclim. CUE")) +
geom_line(mapping= aes(x=control$time,y=warming_VCUE$MIC, color = "Warming + Varying CUE")) +
    labs(x = "Time (yrs)", y = "MIC")


ggarrange(CO2_p, DOC_p, SOC_p, ENZ_p, MIC_p, ncol = 1, common.legend = TRUE, legend = "right")
```

# Step 4: Communicate the results.  

## Methods section

Create a new Rmarkdown document with the following sections.
 
- Conceptual model diagram (you should have this from Lab 7)
- Code is used to simulate the differential equation model as a function.  This is your differential equation function
- Description of the model experiment.

## Result section

Describe results with text and figures.

Since you are using the Rmarkdown, your analysis code will be in the results section.  In your code, specify the parameters (`parms` vector) and the initial conditions (`yini` vector). Run the ODE solver for each of the four simulations.  Then save the output for each simulation as a unique name that you can use to build your graphs. 


## Discussion section

- What did you learn about how microbial dynamics may influence the response of decomposition and soil carbon storage to climate change?  
- Why does the finding from the paper matter?  
- Caveats and key areas of uncertainty in the model. 
- What is the next question that you would like to potentially explore with the model  (or a modified version of the model) and why?   This answer can’t be the same as what is discussed in Allison et al. (2010).

## References

Include any references that you used.  You must include a reference to the Allison et al. (2010) paper.  Use the citation formatting of Ecology.  

**Submit your knitted Rmarkdown that includes your code, figures, and text**



